---
title: "Equivalence Testing for the Structural Interpretation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Equivalence Testing for the Structural Interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `test_t0()` and `test_t1()` functions ask whether the structural
interpretation of a latent factor model can be rejected. They are built to
detect departures, so a small $p$-value is evidence that the indicators do
not depend on the external variable $Z$ only through the latent factor
$\eta$. A large $p$-value is a weaker statement. It does not show that the
structural interpretation holds. It means only that the data did not supply
enough evidence to reject it, which can happen either because the
interpretation is essentially correct or because the sample is too small to
resolve a real departure.

The `equiv_test()` function addresses the complementary question. Instead of
asking whether any departure exists, it asks whether the departure is small
enough to be treated as negligible relative to a margin that the analyst
fixes in advance. A significant result licenses the positive claim that
departures from the structural interpretation lie below the margin, so the
estimates from `fit_structural()` may be read structurally.

This vignette assumes familiarity with the model, the notation, and the
distance metric statistic introduced in the *Getting Started* vignette
(`vignette("structest")`) and the *Standard Errors* vignette
(`vignette("fit-structural")`). We recall the relevant quantities before
defining the equivalence test.

```{r setup}
library(structest)
```

## Recap: The Distance Metric Statistic

Both tests estimate the restricted model by GMM and report the minimised
distance metric statistic

$$
J = Q_N(\hat\theta)
  = N\,\bar U(\hat\theta)^\top\,\Sigma^{-1}\,\bar U(\hat\theta),
$$

where $\bar U(\theta)$ is the sample average of the $d \times p$ moment
conditions and $\Sigma$ is their covariance. The statistic $J$ equals $T_0$
or $T_1$ depending on the parameterisation. Under the null that the
structural interpretation holds, $J$ is asymptotically $\chi^2_{df}$ with
$df = d \times p - m$ overidentifying degrees of freedom, where $m$ is the
number of free parameters. The rejection tests compare $J$ to the upper tail
of this $\chi^2_{df}$ distribution.

## Why a Non-significant Test Is Not Confirmation

The rejection test controls the probability of wrongly rejecting a correct
structural interpretation, but it says nothing direct about the probability
of failing to reject an incorrect one. When the sample is small, or the
departure is subtle, $J$ can easily fall in the non-rejection region even
though the structural interpretation is false. Reporting a large $p$-value as
support for the structural model therefore confuses absence of evidence with
evidence of absence.

The same asymmetry appears from the other side in large samples. There the
rejection test has the power to flag even a negligible departure, so a small
$p$-value confirms that the departure is nonzero without telling us whether it
is large enough to matter. In both regimes the rejection test answers a
different question from the one we usually care about.

Equivalence testing repairs this by inverting the burden of proof. We choose
a margin $\varepsilon_0$ that represents the largest departure we are willing
to call negligible, and we require positive evidence that the true departure
is smaller than $\varepsilon_0$ before we conclude in favour of the
structural interpretation.

## Measuring the Size of a Departure

To speak of a departure being small or large we need a scale. Under a
sequence of local alternatives to the structural model, $J$ is asymptotically
non-central $\chi^2_{df}(\lambda)$. The non-centrality parameter $\lambda$
measures how far the population sits from the restricted model, with
$\lambda = 0$ recovering the central $\chi^2_{df}$ of exact fit.

The non-centrality $\lambda$ is, to first order, $n$ times the population
value of the distance metric, namely the standardised squared distance
between the truth and the closest the restricted model can come. It therefore
grows with sample size and is not directly comparable across studies.
Following the close-fit tradition in covariance structure modelling, we divide
by $n\,df$ to obtain a per-degree-of-freedom average that no longer depends on
sample size, and take the square root to return to the scale of a
standardised discrepancy,

$$
\varepsilon = \sqrt{\frac{\lambda}{n\,df}} .
$$

We call $\varepsilon$ the standardised misfit. It is the root-mean-square
approximation error of the structural model, directly analogous to the RMSEA
in covariance structure modelling. Each of the $df$ overidentifying
restrictions contributes a standardised squared discrepancy, and $\varepsilon$
is the square root of their average. A value of zero corresponds to exact
structural fit, the index is dimensionless because the weighting by
$\Sigma^{-1}$ places every moment on a standardised footing, and larger values
correspond to larger departures. The point estimate used by `equiv_test()`
simply inverts the relationship $E[J] \approx df + \lambda$ and floors it at
zero,

$$
\hat\varepsilon = \sqrt{\max\!\left(0,\ \frac{J - df}{n\,df}\right)}.
$$

## The Close-fit Hypotheses

A margin $\varepsilon_0$ on the $\varepsilon$ scale corresponds to a boundary
non-centrality $\lambda_0 = n\,df\,\varepsilon_0^2$. The equivalence test is a
one-sided close-fit test of

$$
H_0:\ \varepsilon \ge \varepsilon_0
\qquad\text{versus}\qquad
H_1:\ \varepsilon < \varepsilon_0 .
$$

Note the direction. The null is now that the departure is at least as large
as the margin, and rejecting it is what allows the positive conclusion that
the departure is smaller than the margin. This is the opposite arrangement to
`test_t0()` and `test_t1()`, where the null is exact structural fit.

## The Equivalence p-value and the Upper Confidence Limit

Because $J$ is stochastically increasing in $\lambda$, small observed values
of $J$ are evidence against $H_0$. The equivalence $p$-value is the
lower-tail probability of the boundary distribution,

$$
p = P\!\left(\chi^2_{df}(\lambda_0) \le J_{\text{obs}}\right),
$$

and we declare equivalence when $p < \alpha$.

The same decision can be read off a confidence limit. Let
$\varepsilon_{\text{upper}}$ be the one-sided $(1 - \alpha)$ upper confidence
limit for $\varepsilon$, obtained by finding the non-centrality $\lambda_u$
that solves $P(\chi^2_{df}(\lambda_u) \le J_{\text{obs}}) = \alpha$ and
rescaling. The test then has the clean dual form

$$
\text{declare equivalence}
\iff \varepsilon_{\text{upper}} < \varepsilon_0
\iff p < \alpha .
$$

Reporting $\varepsilon_{\text{upper}}$ alongside the decision is often more
informative than the $p$-value, since it states directly how large the
departure could plausibly be.

## Relationship to Two One-sided Tests

Readers familiar with equivalence testing for a mean difference will know the
two one-sided tests (TOST) procedure. There the parameter of interest is a
signed effect $\theta$, and equivalence means $|\theta| < \Delta$. Because
$\theta$ can be either positive or negative, TOST runs two one-sided tests,
one against each bound, and rejects only when both pass. The matching
confidence interval is the $(1 - 2\alpha)$ interval, which for
$\alpha = 0.05$ is the $90\%$ interval rather than the $95\%$ interval.

The structural setting differs in one important respect. The misfit
$\varepsilon$ is a non-negative omnibus quantity, not a signed effect, so
being outside the margin has only one side, namely
$\varepsilon \ge \varepsilon_0$. A single one-sided test is therefore the
correct procedure, and the matching confidence statement is a one-sided
$(1 - \alpha)$ upper limit, not a $(1 - 2\alpha)$ interval. Carrying the TOST
$90\%$ rule over to this setting would be a mistake, because that rule is
specific to the two-bound case.

## Choosing the Margin

The margin is a substantive judgement, not a statistical default, and
`equiv_test()` deliberately has no default value for it. Unlike the
established conventions for the RMSEA in covariance structure modelling, there
is no canonical threshold for what counts as approximate structural
interpretation, so the margin has to be justified on the merits of the
application. A smaller margin is a stronger claim and requires more data to
support. It is good practice to fix the margin before seeing the test result.

## Worked Example

We simulate data with a small direct effect of $Z$ on the first indicator, so
that the structural interpretation is mildly violated, and we use a
moderately large sample. The departure is genuine but modest.

```{r simulate}
set.seed(12345)
n <- 4000
z <- rbinom(n, 1, 0.5)
eta <- 1 + 0.5 * z + rnorm(n)
lambda <- c(1.0, 0.8, 0.6)
X <- cbind(
  2 + lambda[1] * eta + 0.10 * z + rnorm(n, sd = 0.5),  # direct effect of z
  3 + lambda[2] * eta + rnorm(n, sd = 0.5),
  1 + lambda[3] * eta + rnorm(n, sd = 0.5)
)

fit <- test_t0(X, z)
fit
```

At this sample size the rejection test detects the departure and rejects
exact structural fit. That tells us the departure is nonzero, but not whether
it is large. The equivalence test asks the second question. We first take a
margin of $0.08$ on the standardised scale.

```{r equiv}
et <- equiv_test(fit, margin = 0.08)
et
```

The output reports the underlying statistic, the equivalence $p$-value, the
point estimate $\hat\varepsilon$, and the one-sided upper limit
$\varepsilon_{\text{upper}}$. Here the upper limit falls below $0.08$, so the
departure, though real, is small on this scale. A more demanding margin tells
a different story. The same data cannot establish that the misfit is below
$0.05$, because the upper limit exceeds that value.

```{r equiv-strict}
equiv_test(fit, margin = 0.05)
```

Taken together the two analyses say that the structural interpretation is not
exact here, but the departure is modest. It is below $0.08$ on the
standardised scale, and not demonstrably below $0.05$. The conclusion is
always relative to the margin, which is why the margin must be chosen on
substantive grounds before the test is run.

The components are available for programmatic use.

```{r components}
et$epsilon          # point estimate of standardised misfit
et$epsilon_upper    # one-sided (1 - alpha) upper limit
et$equivalence      # decision at the chosen alpha
```

## Calibration

The non-central $\chi^2$ reference is asymptotic, and the mapping from
$\lambda$ to $\varepsilon$ relies on the local-alternatives approximation. In
finite samples the actual size of the test at the boundary need not equal
$\alpha$ exactly, and the approximation is best checked by simulation under a
data-generating process with a known departure. In our own checks the size at
the boundary tracks $\alpha$ closely across a range of sample sizes once the
departure is set so that the population $\varepsilon$ equals the margin, and
$\hat\varepsilon$ is stable across $n$ as the rescaling intends. We recommend
a short calibration study tailored to the design at hand before relying on the
test for a substantive claim.
