---
title: "Fitting the Structural Model with Standard Errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fitting the Structural Model with Standard Errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `test_t0()` and `test_t1()` functions estimate the structural model
parameters via GMM and return both a test statistic and the point estimates
(accessible via `$estimates`). However, they do not provide standard errors
for those estimates---they tell you whether the structural interpretation
can be rejected, but not how precisely each individual parameter is
estimated or whether individual effects are statistically significant.

The `fit_structural()` function fills this gap. It runs the same GMM
estimation as the test functions, then additionally computes sandwich
standard errors for every free parameter. The result is a coefficient
table---analogous to `summary.lm()`---with point estimates, standard
errors, $z$-values, $p$-values, and confidence intervals. The
overidentification test statistic (equivalent to $T_0$ or $T_1$) is
reported as a model diagnostic.

This vignette assumes familiarity with the model and notation introduced in
the *Getting Started* vignette (`vignette("structest")`). We briefly recall
the key definitions below before developing the standard-error machinery.

```{r setup}
library(structest)
```

## Recap: The Structural Model and Moment Conditions

Recall from the *Getting Started* vignette that under the basic latent
factor model (Equation 1 there), each indicator satisfies
$X_i = \lambda_i\,\eta + \varepsilon_i$. The structural interpretation
requires that any external discrete variable $Z$ affects the indicators only
through $\eta$, which imposes testable restrictions on the conditional
means $E(X_i \mid Z = z_j)$.

### $T_1$ parameterisation (reliability-independent)

Under Equation (3) of the *Getting Started* vignette, the restricted model
is

$$
E(X_i \mid Z = z_j) = \gamma_i + \alpha_i\,\beta_j
$$

with identification constraints $\alpha_1 = 1$ and $\beta_1 = 0$. The free
parameter vector is

$$
\theta = (\gamma_1, \ldots, \gamma_d,\; \alpha_2, \ldots, \alpha_d,\;
\beta_2, \ldots, \beta_p)
$$

with $2d + p - 2$ free parameters in total. Here:

- $\gamma_i = E(X_i \mid Z = z_1)$: intercept for indicator $i$ at the
  reference level of $Z$.
- $\alpha_i = \lambda_i / \lambda_1$: the relative loading of indicator $i$
  compared to the first indicator ($\alpha_1 = 1$ is fixed for
  identification).
- $\beta_j$: the shift in the latent factor at $Z$-level $j$ relative to
  the reference ($\beta_1 = 0$).

### $T_0$ parameterisation (reliability-dependent)

Under Equation (2) of the *Getting Started* vignette, the restricted model
is

$$
E(X_i \mid Z = z_j) = \gamma_i +
  \frac{\lambda_i}{\lambda_1}\,\beta_j
$$

with $\beta_1 = 0$ and the ratios $\lambda_i / \lambda_1$ pre-estimated
from pairwise covariances. Here the free parameter vector is

$$
\theta = (\gamma_1, \ldots, \gamma_d,\; \beta_2, \ldots, \beta_p)
$$

with $d + p - 1$ free parameters. The loading ratios are treated as known
constants plugged in from the reliability estimation step.


## Moment Conditions

Both parameterisations are estimated by the generalised method of moments
(GMM). Using the same notation as the *Getting Started* vignette, the
moment conditions are defined by the residuals: for each indicator $i$ and
$Z$-level $j$, observation $k$ contributes

$$
U_{k,(i,j)} = I(Z_k = z_j)\!\left(
  X_{ik} - E_\theta(X_i \mid Z = z_j)\right) \tag{4}
$$

where $E_\theta(\cdot)$ is the predicted conditional mean under the model.
For the $T_1$ parameterisation, this is
$E_\theta(X_i \mid Z = z_j) = \gamma_i + \alpha_i\,\beta_j$; for $T_0$,
it is $E_\theta(X_i \mid Z = z_j) = \gamma_i + (\lambda_i/\lambda_1)\,\beta_j$.

Stacking all $d \times p$ moment conditions gives a vector
$U_k(\theta) \in \mathbb{R}^{d \times p}$. Under correct specification,
$E[U_k(\theta_0)] = 0$ --- the residuals average to zero at the true
parameter values.

### From moments to estimation

The GMM estimator finds the $\hat\theta$ that makes the sample average of
these moment conditions as close to zero as possible. Specifically, it
minimises the distance metric statistic

$$
Q_N(\theta) = N\,\bar U(\theta)^\top\,\Sigma^{-1}\,\bar U(\theta) \tag{5}
$$

where $\bar U(\theta) = N^{-1}\sum_{k=1}^{N} U_k(\theta)$ is the sample
average and $\Sigma$ is the empirical covariance matrix of the $U_k$
(with the variance adjustment for $T_0$ described in the *Getting Started*
vignette). The weight matrix $\Sigma^{-1}$ standardises each moment
condition by its variance and accounts for correlations between them.

### From estimation to testing

At the minimiser $\hat\theta$, the minimised value $Q_N(\hat\theta)$ is
exactly the test statistic $T_0$ or $T_1$. The system has $d \times p$
moment conditions but only $m$ free parameters ($m = 2d + p - 2$ for $T_1$;
$m = d + p - 1$ for $T_0$). The excess $d \times p - m$ conditions provide
the overidentification test: if the model is correct, $Q_N(\hat\theta)$
is asymptotically $\chi^2$ with $d \times p - m$ degrees of freedom.

### From estimation to standard errors

The same GMM framework also provides standard errors for $\hat\theta$.
These come from asking: how sensitive is $\hat\theta$ to perturbations in
the data? The answer involves two quantities:

- **The Jacobian** $G$: how the moment conditions $E[U_k]$ respond to
  changes in $\theta$. This captures the "curvature" of the objective
  function.
- **The variance** $\Sigma$: how noisy the individual moment contributions
  $U_k$ are.

The next section derives the sandwich variance formula from these two
ingredients.


## Sandwich Standard Errors

### The general sandwich formula

For a GMM estimator with weight matrix $W$, the asymptotic variance is
(Newey & McFadden, 1994, Theorem 3.4):

$$
\mathrm{Var}(\hat\theta) = \frac{1}{N}
  \left(G^\top W G\right)^{-1}
  G^\top W\,\Sigma\,W G
  \left(G^\top W G\right)^{-1}
$$

where $G = E\!\left[\partial U_k(\theta) / \partial \theta^\top\right]$
is the $q \times m$ Jacobian ($q = d \cdot p$ moments, $m$ free parameters)
and $\Sigma = \mathrm{Var}(U_k)$ is the $q \times q$ variance of
per-observation moment contributions.

### Simplification under efficient GMM

Our estimator uses the efficient weight matrix $W = \Sigma^{-1}$ (Hansen,
1982). Substituting into the general formula, the middle term collapses:

$$
(G^\top \Sigma^{-1} G)^{-1}\,
G^\top \underbrace{\Sigma^{-1}\,\Sigma\,\Sigma^{-1}}_{\Sigma^{-1}}\,G\,
(G^\top \Sigma^{-1} G)^{-1}
= (G^\top \Sigma^{-1} G)^{-1}
$$

giving the efficient sandwich variance:

$$
\mathrm{Var}(\hat\theta) = \frac{1}{N}
  \left(G^\top\,\Sigma^{-1}\,G\right)^{-1} \tag{6}
$$

This is what `fit_structural()` computes. The same sandwich formula is used
in the broader SEM literature---for instance, Bollen, Kolenikov, & Bauldry
(2014) apply it to general latent variable models via model-implied
instrumental variables. Our setting is simpler: the moment conditions come
directly from the restricted conditional means, so $G$ has a closed-form
expression and no instrumental variable selection is required.

The key computational task is deriving $G$ analytically and estimating
$\Sigma$ from the data.

### Jacobian for the $T_1$ model

Under the $T_1$ parameterisation, the moment condition for indicator $i$
and $Z$-level $j$ at observation $k$ is

$$
U_{k,(i,j)} = I(Z_k = z_j)\!\left(
  X_{ik} - \gamma_i - \alpha_i\,\beta_j\right).
$$

The free parameter vector is
$\theta = (\gamma_1, \ldots, \gamma_d,\; \alpha_2, \ldots, \alpha_d,\;
\beta_2, \ldots, \beta_p)$, with $\alpha_1 = 1$ and $\beta_1 = 0$ fixed.
The Jacobian $G$ has entries
$G_{(i,j),\,\ell} = E[\partial U_{k,(i,j)} / \partial \theta_\ell]$. Let
$\pi_j = P(Z = z_j)$.

**Derivative with respect to $\gamma_s$** ($s = 1, \ldots, d$):

$$
E\!\left[\frac{\partial U_{k,(i,j)}}{\partial\,\gamma_s}\right]
  = -\pi_j\,\mathbf{1}(i = s).
$$

Only the moment conditions for indicator $s$ are affected.

**Derivative with respect to $\alpha_s$** ($s = 2, \ldots, d$):

$$
E\!\left[\frac{\partial U_{k,(i,j)}}{\partial\,\alpha_s}\right]
  = -\pi_j\,\beta_j\,\mathbf{1}(i = s).
$$

This involves the product with $\beta_j$: when the $Z$-effect $\beta_j$ is
near zero, the moment conditions carry little information about $\alpha_s$,
and its standard error will be large.

**Derivative with respect to $\beta_t$** ($t = 2, \ldots, p$):

$$
E\!\left[\frac{\partial U_{k,(i,j)}}{\partial\,\beta_t}\right]
  = -\pi_j\,\alpha_i\,\mathbf{1}(j = t).
$$

Indicators with larger loadings ($\alpha_i$) contribute more information
about the $Z$-effect.

**Assembling $G$.**
The Jacobian is a $(d \cdot p) \times (2d + p - 2)$ matrix. Row $(i, j)$
corresponds to the moment condition for indicator $i$ at $Z$-level $j$.
In practice, $\pi_j$ is estimated by the sample proportion
$\hat\pi_j = n_j / N$.

### Jacobian for the $T_0$ model

Under the $T_0$ parameterisation, the moment condition is

$$
U_{k,(i,j)} = I(Z_k = z_j)\!\left(
  X_{ik} - \gamma_i - r_i\,\beta_j\right)
$$

where $r_i = \lambda_i / \lambda_1$ is the pre-estimated loading ratio,
treated as a known constant. The free parameter vector is
$\theta = (\gamma_1, \ldots, \gamma_d,\; \beta_2, \ldots, \beta_p)$.

**Derivative with respect to $\gamma_s$:**
Identical to the $T_1$ case: $-\pi_j\,\mathbf{1}(i = s)$.

**Derivative with respect to $\beta_t$** ($t = 2, \ldots, p$):

$$
E\!\left[\frac{\partial U_{k,(i,j)}}{\partial\,\beta_t}\right]
  = -\pi_j\,r_i\,\mathbf{1}(j = t).
$$

The Jacobian is a $(d \cdot p) \times (d + p - 1)$ matrix---smaller than
the $T_1$ Jacobian because there are no $\alpha$ parameters. Since
$r_i$ is treated as fixed, the Jacobian does not capture uncertainty in the
reliability estimates; that uncertainty is instead absorbed into $\Sigma$
through the variance adjustment described in the *Getting Started* vignette
(see *Variance adjustment for estimated reliabilities*).


## Example: Fitting the $T_1$ Model

```{r fit-t1}
set.seed(12345)
n <- 1000

# Simulate under the structural model
z <- sample(0:2, n, replace = TRUE)
eta <- 1 + 0.3 * (z == 1) + 0.7 * (z == 2) + rnorm(n)
lambda <- c(1.0, 0.8, 0.6)
X <- cbind(
  2 + lambda[1] * eta + rnorm(n, sd = 0.5),
  3 + lambda[2] * eta + rnorm(n, sd = 0.5),
  1 + lambda[3] * eta + rnorm(n, sd = 0.5)
)

fit1 <- fit_structural(X, z, type = "t1")
summary(fit1)
```

The coefficient table reports:

- **gamma**: the intercept for each indicator, equal to $E(X_i \mid Z = z_1)$
  at the reference level. These combine the indicator mean $\mu_i$ and the
  baseline latent factor level $\lambda_i E(\eta \mid Z = z_1)$.
- **alpha**: the relative loading of each indicator, with $\alpha_1 = 1$
  fixed for identification. Under the structural model,
  $\alpha_i = \lambda_i / \lambda_1$.
- **beta**: the shift in the latent factor at each $Z$-level relative to
  the reference, with $\beta_1 = 0$.

The overidentification test (J statistic) is equivalent to the $T_1$ test
from `test_t1()`. A large $p$-value indicates that the structural model
fits the data adequately.


## Example: Fitting the $T_0$ Model

```{r fit-t0}
fit0 <- fit_structural(X, z, type = "t0")
summary(fit0)
```

The $T_0$ model estimates fewer parameters (no $\alpha_i$), since the loading
ratios are pre-estimated from pairwise covariances. The estimated
reliabilities $\hat\lambda$ are reported in the summary output. The standard
errors account for the uncertainty in these reliability estimates.


## Extracting Results

The `structest_fit` object supports standard R model-object methods:

```{r methods}
# Point estimates (named vector of free parameters)
coef(fit1)

# Confidence intervals (default: 95%)
confint(fit1)

# 90% confidence intervals
confint(fit1, level = 0.90)

# Variance-covariance matrix
vcov(fit1)
```

The `confint()` method accepts a `parm` argument for selecting specific
parameters:

```{r confint-parm}
# CI for beta parameters only
confint(fit1, parm = c("beta[z=1]", "beta[z=2]"))
```


## Interpreting the Parameters

The parameter estimates from `fit_structural()` are meaningful **only when
the structural model is not rejected**. When the overidentification test
yields a large $p$-value, the model fits the data adequately and the
parameters describe the latent factor structure:

- **$\beta_j$** measures how much the latent factor $\eta$ shifts at
  $Z$-level $j$ relative to the reference. A significant $\beta_j$ means
  $Z$-level $j$ is associated with a different level of the latent factor.
- **$\alpha_i$** ($T_1$ only) gives the relative loading of indicator $i$
  compared to the reference indicator ($\alpha_1 = 1$). Under the structural
  model, $\alpha_i = \lambda_i / \lambda_1$.
- **$\gamma_i$** is the intercept for indicator $i$ at the reference level
  of $Z$: $\gamma_i = \mu_i + \lambda_i\,E(\eta \mid Z = z_1)$.

### When the model is rejected

If the overidentification test rejects (small $p$-value), the structural
model does not hold and the parameter estimates should not be interpreted
substantively. Under misspecification, the model is fitting a rank-1
structure to data that does not have one, so the estimates do not correspond
to well-defined population quantities and the standard errors are not valid.

In particular, the per-parameter $p$-values **cannot** be used to diagnose
*which* indicator breaks the structural interpretation. The estimates and
their standard errors are derived under the assumption that the model is
correct; when it is not, they do not have a meaningful interpretation.


## Relationship to the Tests

The `fit_structural()` function and the `test_t0()`/`test_t1()` functions
share the same underlying GMM estimation. The point estimates are identical:

```{r compare}
t1_test <- test_t1(X, z)
fit1 <- fit_structural(X, z, type = "t1")

# Same point estimates
all.equal(t1_test$estimates$gamma, fit1$estimates$gamma)
all.equal(t1_test$estimates$alpha, fit1$estimates$alpha)
all.equal(t1_test$estimates$beta, fit1$estimates$beta)

# Same test statistic
all.equal(unname(t1_test$statistic), fit1$test$statistic)
```

The point estimates and test statistic are already available from the test
functions. The difference is that `fit_structural()` additionally computes
the Jacobian $G$ and the sandwich variance (Equation 6), providing standard
errors, confidence intervals, and per-parameter significance tests that the
test functions do not.


## References

Bollen, K. A., Kolenikov, S., & Bauldry, S. (2014). Model-implied
instrumental variable-generalized method of moments (MIIV-GMM) estimators
for latent variable models. *Psychometrika*, *79*(1), 20--50.
https://doi.org/10.1007/s11336-013-9335-3

Hansen, L. P. (1982). Large sample properties of generalized method of
moments estimators. *Econometrica*, *50*(4), 1029--1054.
https://doi.org/10.2307/1912775

Newey, W. K., & McFadden, D. (1994). Large sample estimation and hypothesis
testing. In R. F. Engle & D. L. McFadden (Eds.), *Handbook of Econometrics*
(Vol. 4, pp. 2111--2245). Elsevier.
https://doi.org/10.1016/S1573-4412(05)80005-4

VanderWeele, T. J., & Vansteelandt, S. (2022). A statistical test to reject
the structural interpretation of a latent factor model. *Journal of the Royal
Statistical Society: Series B (Statistical Methodology)*, *84*(5), 2032--2054.
https://doi.org/10.1111/rssb.12555
